```{r, echo=FALSE, purl=FALSE, message = FALSE}

library(sotu)
library(tidyverse)
library(tidytext)
library(readtext)

knitr::opts_chunk$set(results='hide', comment = "#>", purl = FALSE)
```

# Analysing Texts {#textanalysis}


> Learning Objectives
>
> - to come

------------

We'll use several libraries today. `sotu` will provide the metadata and text of State of the Union speeches ranging from George Washington to Barack Obama. `tidyverse` provides many of the standard "verbs" for working with our data. `tidytext` provides specific functions for a "tidy" approach to working with textual data. `readtext` provides a function well suited to reading textual data from a large number of formats into R.  

```{r load-libs, eval=FALSE}
library(sotu)
library(tidyverse)
library(tidytext)
library(readtext)
```


## Reading text into R

First, let's look at the data in the `sotu` package. The metadata and texts come separately. We'll use the supplied metadata object, but we're going to use a utility function (`sotu_dir`) in the package to write the texts to disk so that we can practice reading text files from disk. 

```{r}
# Let's take a quick look at the state of the union metadata
summary(sotu_meta)
# sotu_dir writes the text files to a temporary dir, but you could specific where you want them.
fp <- sotu_dir()
head(fp)
```

Now that we have the files in disk, and a list of filepaths stored in the `fp` variable, we can use `readtext` to read the texts into a new variable.

```{r}
# let's then read in the files with readtext
texts <- readtext(fp)
head(texts)
```

So that we can work with a single tabular dataset with a tidy approach, we'll convert the metadata and text tables to tibbles, and combine them into a single tibble. You can see that our texts are organized by alphabetical order, so first we'll need to sort our metadata to match. 

```{r}
sotu_meta_tib <- as_tibble(sotu_meta) %>%
  arrange(president)

head(sotu_meta_tib)
```


```{r}
# We can now combine the sotu metadata with the texts
# first, we'll turn both pieces of data into tibbles, then combine
sotu_texts <- as_tibble(texts)
sotu_whole <- bind_cols(sotu_meta_tib, sotu_texts)
glimpse(sotu_whole)
```

Now that we have our data, we need to think about cleaning it. Depending on the quality of your data, you might need to explicitly replace certain characters or words, remove urls or types of numbers, such as phone numbers, or otherwise clean up misspellings or errors. There are several ways to handle this sort of cleaning, but we'll look at some straightforward string manipulation and replacement. 


## String operations

TODO

(For spell check: https://CRAN.R-project.org/package=spelling,  https://CRAN.R-project.org/package=hunspell)

## Tokenize, lowercase

A very common part of data cleaning involves tokenization. While our data is already "tidy" insofar as each row is a single observation, a single text with metdata, the tidytext approach goes a step further to make each word it's own observation with metadata. We could write our own function to do this using a tokenizer, but `tidytext` provides a handy utility function just for this purpose.

```{r}
tidy_sotu <- sotu_whole %>%
  unnest_tokens(word, text)

tidy_sotu
```

Before we move on, we should note that the `unnest_tokens` function didn't just tokenize our texts at the word level. It also lowercased each word, and it could do quite a bit more. For instance, we could tokenize the text at the level of ngrams or sentences, if those are the best units of analysis for our work. We could also leave punctuation, which has been removed by default. Depending on what you need to do for analysis, you use do these operations during this step, or write custom functions and do it before you unnest tokens.


## Stopwords 

Another common type of cleaning in text analysis is to remove stopwords, or common words that theoretically provide less information about the content of a text. Depending on the type of analysis you're doing, you might leave these words in or use a highly curated list of stopwords. For now, as we move toward looking at words in documents based on frequency, we will remove some standard stopwords using a tidytext approach. 

First, let's look at the stopwords that tidytext gives us to get a sense of what they are.

```{r}
data(stop_words)
head(stop_words, n = 60)
```

You can see that we now have one word per row with associated metadata. We can now remove stopwords using an `anti-join`.

```{r}
tidy_sotu_words <- tidy_sotu %>% 
  anti_join(stop_words)

tidy_sotu_words
```

We went from 1,965,212 to 778,161 rows, which means we had a lot of stopwords in our corpus. This is a huge removal, so for serious analysis, we might want to take a closer look at the stopwords and determine if we should use a different stopword list or otherwise create our own. 


```{r}
# Word tokenization with punctuation
tidy_sotu_w_punct <- sotu_whole %>%
  unnest_tokens(word, text, strip_punct = FALSE)

tidy_sotu_w_punct

# Sentence tokenization
tidy_sotu_sentences <- sotu_whole %>%
  unnest_tokens(sentence, text, token = "sentences", to_lower = FALSE)

tidy_sotu_sentences

# N-gram tokenization
tidy_sotu_trigram <- sotu_whole %>%
  unnest_tokens(trigram, text, token = "ngrams", n = 3)

tidy_sotu_trigram
```
 

## Stemming and Lemmatization? 

- hunspell, SnowballC?


## Tagging

TODO?: Tag text with cleanNLP maybe?
 
Now that we've read in our text and metadata, reshaped it a bit into the tidytext format, and cleaned it a bit while doing so, let's move on to some basic analysis. 

