```{r, echo=FALSE, purl=FALSE, message = FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE,comment = "#>", purl = FALSE)
```

# Analyzing Texts {#textanalysis}


> Learning Objectives
>
> - perform different frequency counts and generate plots
> - use the `widyr` package to calculate co-ocurrance
> - use `igraph` and `ggraph` to plot a co-ocurrance graph
> - import and export a Document-Term Matrix into `tidytext`
> - use the `sentiments` dataset from `tidytext` to perform a sentiment analysis


------------


Now that we've read in our text and metadata, tokenized and cleaned it a little, let's move on to some analysis. 

First, we'll make sure we have loaded the libraries we'll need. 

```{r load-tidylibs, eval=FALSE}
library(tidyverse)
library(tidytext)
```

Let's remind ourselves of what our data looks like. 

```{r}
tidy_sotu_words
```

## Frequencies

Since our unit of analysis at this point is a word, let's do some straightforward counting to figure out which words occur most frequently in the corpus as a whole. 

```{r word-freq}
tidy_sotu_words %>%
  count(word, sort = TRUE)
```

We can pipe this into `ggplot` to make a graph of the words that occur more that 2000 times. We count the words and use `geom_col` to represent the n values.

```{r word-freq-plot}
tidy_sotu_words %>%
  count(word) %>% 
  filter(n > 2000) %>% 
  mutate(word = reorder(word, n)) %>%  # reorder values by frequency
  ggplot(aes(word, n)) +
     geom_col(fill = "gray") +
     coord_flip()  # flip x and y coordinates so we can read the words better
```


What if we want to check the most common words per speech for a single president and see which of the top words apppear in which speech?

```{r obama-topwords}
tidy_sotu_words %>%
  filter(president == "Barack Obama") %>%
  count(doc_id, word) %>%
  filter(n > 20) %>%
  mutate(word = reorder(word, n)) %>% 
  ggplot(aes(word, n, fill=doc_id)) +
    geom_col() +
    coord_flip()
```


>>> CHALLENGE: In any given year, how often is the word 'peace' used and how often is the word 'war' used? Make a bar chart that shows for each year the proportion of each of these two words out of the total of how often both words are used. Bonus: use decades instead of years.
 
```{r challenge-word-years, eval=FALSE, echo=FALSE}
# steps:
# Select only the words 'war' and 'peace'.
# count ocurrences of each per year
# plot n by year, and use position 'fill' to show the proportion

tidy_sotu_words %>%
  filter(word %in% c("war", "peace")) %>% 
  count(year, word) %>% 
  ggplot(aes(year, n, fill = word)) +
    geom_col(position = "fill")

# with decade instead of year
tidy_sotu_words %>%
  filter(word %in% c("war", "peace")) %>% 
  mutate(decade = floor(year/10) * 10) %>% 
  count(decade, word) %>% 
  ggplot(aes(decade, n, fill = word)) +
    geom_col(position = "fill")
```


As another example let us calculate the average number of words per speech for each president. How long was the average speech of each president and who are the top 'wordiest' presidents?  

First we summarize the words per president per speech

```{r word-president-count}
tidy_sotu_words %>%
  count(president, doc_id)
```

Then we use the output table and group it by president. That allows us to calculate the average number of words per speech.

```{r word-president-avg}
tidy_sotu_words %>%
  count(president, doc_id)  %>% 
  group_by(president) %>% 
  summarize(avg_words = mean(n)) %>% 
  arrange(desc(avg_words))
```


## Term frequency

Often a raw count of a word is less important than understanding how often that word appears relative to the total number of words in a text. This ratio would be the **term frequency**. We can use `dplyr` to calculate it like this:

```{r termfreq}
tidy_sotu_words %>%
  count(doc_id, word, sort = T)  %>%  # count n for each word
  group_by(doc_id) %>% 
  mutate(n_tot = sum(n), # count total number of words per doc
         term_freq = n/n_tot)
```

Let's graph the term frequency for one of these speeches so we can understand the frequency distribution of words over a text.

```{r termfreq-plot, message=FALSE}
tidy_sotu_words %>%
  filter(doc_id == "harry-s-truman-1946.txt") %>%
  count(doc_id, word)  %>%  # count n for each word
  group_by(doc_id) %>% 
  mutate(n_tot = sum(n), # count total number of words per doc
         term_freq = n/n_tot) %>% 
  ggplot(aes(term_freq)) +
    geom_histogram() 
```

This distribution makes sense. Most words are used relatively rarely in a text. Only a few have a high term frequency. 

We could keep filtering this data to see which terms have high frequency, thus maybe increased significance, for different presidents and different particular speeches. 

>>> CHALLENGE: Pick one president. For each of his speeches, which is the term with highest term frequency? Create a table as output. (Hint: `top_n`might be useful)

```{r challenge-termfreq, eval=FALSE, echo=FALSE}
tidy_sotu_words %>%
  filter(president == "Ronald Reagan") %>% 
  count(doc_id, word)  %>%  
  group_by(doc_id) %>% 
  mutate(n_tot = sum(n), 
         term_freq = n/n_tot) %>% 
  top_n(1)
```


## Tf-idf 

So far we've been looking at term frequency per document. What if we want to know about words that seem more important based on the contents of the entire corpus?

For this, we can use **term-frequency according to inverse document frequency**, also callled **tf-idf**. Tf-idf measures how important a word is within a corpus by scaling term frequency per document according to the inverse of the term's document frequency (number of documents within the corpus in which the term appears divided by the number of documents). 

The tf-idf value will be:

- lower for words that appear in many documents in the corpus, and lowest when the word occurs in virtually all documents. 
- high for words that appear many times in few documents in the corpus, this lending high discriminatory power to those documents. 

The `tidytext` package includes a function `bind_tf_idf`. It takes a table that contains one-row-per-term-per-document, the name of the column that contains the words (terms), the name of the column which contains the doc-id, and the name of the column that contains the document-term counts.

So below we aggregate our tibble with the word tokens to create the one-row-per-term-per-document table and then pipe it into the `bind_tf_idf` function.

```{r tf-idf}
tidy_sotu_words %>%
  count(doc_id, word, sort = TRUE)  %>%  # aggregate to count n for each word
  bind_tf_idf(word, doc_id, n) 
```

We can see in the output that our function added three columns to our aggregated table: which contain term frequency, inverse document frequency and Tf-idf. 


Let's look at some of the words in the corpus that have the highest tf-idf scores, which means words that are particularly distinctive for their documents. 

```{r tf-idf-sort}
tidy_sotu_words %>%
  count(doc_id, word, sort = TRUE)  %>% 
  bind_tf_idf(word, doc_id, n) %>% 
  arrange(desc(tf_idf))
```

To understand the occurrence of the years as being particularly distinctive we might need to look more closely at the speeches themselves, and determine whether the years are significant or whether they need to be removed from the text either permanently in the clean up or temporarily using `filter`.


>>> Challenge: Pick the same president you chose above. For each of his speeches, which is the term with highest tf-idf? Create a table as output. (Hint: Remember to group by doc_id before you use top_n)

```{r challenge-tfidf, eval=FALSE, echo=FALSE}
tidy_sotu_words %>%
  filter(president == "Ronald Reagan") %>%
  count(doc_id, word) %>%
  bind_tf_idf(word, doc_id, n) %>%
  group_by(doc_id) %>% 
  top_n(1)
```


## N-Grams

We mentioned n-grams in the intro, but let's revisit them here and take a look at the most common bigrams in the speeches. Remember we use the `unnest_token` function on our texts and explicitly tell it to generate bigrams:

```{r bigrams-unnest}
sotu_whole %>%
  unnest_tokens(bigram, text, token = "ngrams", n = 2) # create bigram
```

Let's see the most common bigrams:

```{r bigrams-count}
sotu_whole %>%
  unnest_tokens(bigram, text, token = "ngrams", n = 2) %>% 
  count(bigram, sort = TRUE) # count ocurrences and sort descending
```

Ok, so we again need to remove the stopwords. First let us separate the two words into two columns "word1" and "word2" with `separate` from the `tidyr` package:

```{r bigrams-sep-words}
sotu_whole %>%
  unnest_tokens(bigram, text, token = "ngrams", n = 2) %>% 
  separate(bigram, c("word1", "word2"), sep = " ")
```

Now we use dplyr's `filter` function to select only the words in each column that are not in the stopwords.

```{r bigrams-remove-stop}
sotu_whole %>%
  unnest_tokens(bigram, text, token = "ngrams", n = 2) %>% 
  separate(bigram, c("word1", "word2"), sep = " ") %>% # separate into cols
  filter(!word1 %in% stop_words$word) %>% # remove stopwords
  filter(!word2 %in% stop_words$word)
```

Lastly, we re-unite the two word columns into back into our bigrams and save it into a new table `sotu_bigrams`.

```{r bigrams-unite}
sotu_bigrams <- sotu_whole %>%
  unnest_tokens(bigram, text, token = "ngrams", n = 2) %>% 
  separate(bigram, c("word1", "word2"), sep = " ") %>% # separate into cols
  filter(!word1 %in% stop_words$word) %>% # remove stopwords
  filter(!word2 %in% stop_words$word) %>% 
  unite(bigram, word1, word2, sep = " ")  # combine columns


sotu_bigrams %>% 
  count(bigram, sort = TRUE)
```

(Bonus question: What happened on that June 30th?)

A bigram can also be treated as a term in a document in the same way that we treated individual words. That means we can look at tf-idf values in the same way. For example, we can find out the most distinct bigrams that the presidents uttered in all their respective speeches taken together.

We group by president and bigram and then bind the tf-idf value with the `bind_tf_idf` function.

```{r bigram-tf-idf}
sotu_bigrams %>%
  count(president, bigram) %>%
  bind_tf_idf(bigram, president, n) %>%
  arrange(desc(tf_idf))
```

>>> Challenge: Again, pick the same president you chose above. For each of his speeches, which is the bigram with highest tf-idf? Create a table as output. 

```{r challenge-bigram-tfidf, eval=FALSE, echo=FALSE}
sotu_bigrams %>%
  filter(president == "Ronald Reagan") %>% 
  count(doc_id, bigram) %>%  
  bind_tf_idf(bigram, doc_id, n) %>%
  group_by(doc_id) %>% 
  top_n(1)
```



## Co-occurrence

Co-occurrences give us a sense of words that appear in the same text, but not necessarily next to each other.

For this section we will make use of the `widyr` package. The function which helps us do this is the `pairwise_count()` function. It lets us count common pairs of words co-appearing within the same speech.

Behind the scenes, this function first turns our table into a wide matrix. In our case that matrix will be made up of the individual words and the cell values will be the counts of in how many speeches they co-occur, like this:

    | we  | thus | have 
we |  NA  |  4   | 5
thus | 4  | NA   | 2
have  | 5 | 2    | NA


It then will turn the matrix back into a tidy form, where each row contains the word pairs and the count of their co-occurrence. Since we don't care about the order of the words, we will not count the upper triangle of the wide matrix, which leaves us with:

we | thus | 4
we | have | 5
thus | have | 2

Since processing the entire corpus would take too long here, we will only look at the last 100 words of each speech: which words occurr most commonly together at the end of the speeches?

```{r pairwise-count}
library(widyr)

sotu_word_pairs <- sotu_whole %>% 
  mutate(speech_end = word(text, -100, end = -1)) %>%  # extract last 100 words
  unnest_tokens(word, speech_end) %>%   # tokenize
  filter(!word %in% stop_words$word) %>%  # remove stopwords
  pairwise_count(word, doc_id, sort = TRUE, upper = FALSE) # don't include upper triangle of matrix

sotu_word_pairs
```

To visualize the co-occurrence network of words that occur together at the end of 10 or more speeches, we use the `igraph` package to convert our table into a network graph and the `ggraph` package which adds functionality to ggplot to make it easier to plot a network.

```{r plot-network} 
library(igraph)
library(ggraph)

sotu_word_pairs %>% 
  filter(n >= 10) %>%  # only word pairs that occur 10 or more times
  graph_from_data_frame() %>% #convert to graph
  ggraph(layout = "fr") + # place nodes according to the force-directed algorithm of Fruchterman and Reingold
  geom_edge_link(aes(edge_alpha = n, edge_width = n), edge_colour = "tomato") +
  geom_node_point(size = 5) +
  geom_node_text(aes(label = name), repel = TRUE, 
                 point.padding = unit(0.2, "lines")) +
  theme_void()
```

There are alternative approaches for this as well. See for example the `findAssocs` function in the `tm` package.


## Document-Term Matrix

A [document-term matrix (DTM)](https://en.wikipedia.org/wiki/Document-term_matrix) is a format which is frequently used in text analysis. It is a matrix where we can see the counts of each term per document. In a DTM each row represents a document, each column represents a term, and the cell values are the counts of the occurrences of the term for the particular document.

`tidytext` provides functionality to convert to and from DTMs, if for example, your analyis requires specific functions that require you to use a different R package which only works with DTM objects.

The `cast_dtm` function can be used to create a DTM object from a tidy table.

Let's assume that for some reason we want to use the `findAssoc` function from the `tm` package.

First we use dplyr to create a table with the document name, the term, and the count.

```{r term-count}
# make a table with document, term, count
tidy_sotu_words %>% 
  count(doc_id, word) 
```

Now we cast it as a DTM.

```{r cast-dtm}
sotu_dtm <- tidy_sotu_words %>% 
  count(doc_id, word) %>% 
  cast_dtm(doc_id, word, n) 

class(sotu_dtm)
```

Finally, let's use it in the `tm` package.

```{r tm-dtm}
library(tm)

# look at the terms with tm function
Terms(sotu_dtm) %>% tail()

# most frequent terms
findFreqTerms(sotu_dtm, lowfreq = 5000)
              
# find terms associated with ...
findAssocs(sotu_dtm, "citizen", corlimit = 0.5)
```

Conversely, `tidytext` implements the `tidy` function (originally from the `broom` package) to import DocumentTermMatrix objects. Note that it only takes the cells from the DTM that are not 0, so there will be no rows with 0 counts.


## Sentiment analysis

`tidytext` comes with a dataset `sentiments` which contains several sentiment lexicons, where each word is attributed a certain sentiment, like this:

```{r sentiments}
sentiments
```

Here we will take a look at how the sentiment of the speeches change over time. We will use the lexicon from [Bing Liu and collaborators](https://www.cs.uic.edu/~liub/FBS/sentiment-analysis.html), which assigns positive/negative labels for each word:

```{r bing}
bing_lex <- get_sentiments("bing")
bing_lex
```

We can use these sentiments attached to each word and join them to the words of our speeches. We will use `inner_join` from `dplyr`. It will take all rows with words from `tidy_sotu_words` that match words in `bing_lex`, eliminating rows where the word cannot be found  in the lexicon. Since our columns to join on have the same name (`word`) we don't need to explicitly name it.

```{r join-sentiments}
sotu_sentiments <- tidy_sotu_words %>% 
  inner_join(bing_lex)  # join to add semtinemt column

sotu_sentiments
```


Finally we can visualize the proportion of positive sentiment out of the total of positive and negative over time like this:

```{r sentiment-plot, fig.width = 6, fig.height = 4}
sotu_sentiments %>% 
  count(year, sentiment) %>% # group by year and sentiment
  spread(sentiment, n) %>% # spread by seniment
  mutate(pct_positive_sentiment = positive/(negative + positive) * 100) %>% # calculate %
  ggplot(aes(year, pct_positive_sentiment)) +
    geom_line(color="gray") +
    geom_smooth(span = 0.3, se = FALSE) + # for easier viewing
    geom_hline(yintercept = 50, linetype="dotted") + # 50% as reference
    scale_x_continuous(breaks = seq(1790, 2016, by = 10)) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

