```{r, echo=FALSE, purl=FALSE, message = FALSE}
knitr::opts_chunk$set(results='hide', comment = "#>", purl = FALSE)
```

# Preparing Textual Data {#textprep}


> Learning Objectives
>
> - to come

------------

First, we'll load the libraries we need. 

```{r load-tidylibs, eval=FALSE}
library(tidyverse)
library(tidytext)
```

Let's remind ourselves of what our data looks like. 

```{r}
tidy_sotu_words
```

## Frequencies

Since our unit of analysis at this point is a word, let's do some straightforward counting to figure out which words occur most frequently in the corpus as a whole. 

```{r}
tidy_sotu_words %>%
  count(word, sort = TRUE)
```

We could start adding in a bit of visualization here. Let's show the most frequent words that occur more than 2000 times. 

```{r}
tidy_sotu_words %>%
  count(word, sort = TRUE) %>%
  filter(n > 2000) %>%
  mutate(word = reorder(word, n)) %>%
  ggplot(aes(word, n)) +
  geom_col() +
  xlab(NULL) +
  coord_flip()

```

What if we're interested in most used words per speech?

```{r}
# Count words by book
doc_words <- tidy_sotu_words %>%
  count(doc_id, word, sort = TRUE)

# Calculate the total number of words by book and save them to a tibble
total_words <- doc_words %>%
  group_by(doc_id) %>%
  summarize(total = sum(n))

# Join the total column with the rest of the data so we can calculate frequency
doc_words <- left_join(doc_words, total_words)

doc_words 
```

Let's graph the top words per book

```{r}
doc_words %>% 
  filter(n > 100) %>%
  ggplot(aes(word, n, fill = doc_id)) +
  geom_col() + 
  xlab(NULL) +
  coord_flip()
```

That's cool looking, but let's split it into facets so we can see by speech. 

```{r}
doc_words %>% 
  filter(n > 100) %>%
  ggplot(aes(word, n, fill = doc_id)) +
  geom_col(show.legend = FALSE) + 
  xlab(NULL) +
  facet_wrap(~doc_id, ncol = 2) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

We could keep cleaning this figure up by setting some minimum sizing, determining the spacing between y-axis labels better, and so forth, but now we'll accept it as showing some sense of variation across speeches where certain words are used most. 

What if we want to check the most highly common words per speech for a single president? We could filter this `doc_words` dataset based on the president's name being in the doc_id, but I think it's easier to filter from the initial tidy data and recount. 

```{r}
tidy_sotu_words %>%
  filter(president == "Barack Obama") %>%
  count(doc_id, word, sort = TRUE) %>%
  filter(n > 20) %>%
  ggplot(aes(word, n, fill=doc_id)) +
  geom_col() +
  facet_wrap(~doc_id, ncol = 2) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


## Term frequency

Sometimes, a raw count of a word is less important than understanding how often that word appears in respect to the total number of words in a text. This ratio would be the **term frequency**. 

```{r}
doc_words <- doc_words %>%
  mutate(term_freq = n / total)

doc_words 
```

Let's graph the term frequency for one of these speeches so we can understand the frequency distribution of words over a text.

```{r}
doc_words %>%
  filter(doc_id == "harry-s-truman-1946.txt") %>%
  ggplot(aes(term_freq)) +
  geom_histogram(show.legend = FALSE) +
  xlim(NA, .012)
```

This should make sense. Most words are used relatively rarely in a text. Only a few have a high term frequency. 

We could keep filtering this data to see which terms have the high frequency, thus maybe increased significance, for different presidents and different particular speeches. We could also subset based on decade, and get a sense of what was important in each decade. We're going to take a slightly different approach though. We've been looking at term frequency per document. What if we want to know about words that seem more important based on the contents of the entire corpus?


## Tf-idf 

For this, we can use term-frequency according to inverse document frequency (tf-idf). Tf-idf meansures how important a word is within a corpus by scaling term frequency per document according to the inverse of the term's document frequency (how many documents within the corpus in which the term appears divided by the number of documents). 

We could write our own function for tf-idf, but in this case we'll take advantage of tidytext's implementation. 

```{r}
doc_words <- doc_words %>%
  bind_tf_idf(word, doc_id, n)

doc_words
```

The tf-idf value will be:

- lower for words that appear in many documents in the corpus, and lowest when the word occurs in virtually all documents. 
- high for words that appear many times in few documents in the corpus, this lending high discrimiatory power to those doucments. 

Let's look at some of the words in the corpus that have the highest tf-idf scores, which means words that are particularly distinctive for their documents. 

```{r}
doc_words %>%
  select(-total) %>%
  arrange(desc(tf_idf))
```

These results seem appropriate given our history. To understand the occurence of the years we might need to look more closely at the speeches themselves, and determine whether the years are significant or whether they need to be removed from the text. It might be that even if they don't need to be removed from the text overall, they still need to be filtered out within the context of this analysis. 

In the same way that we narrowed our analysis to Obama speeches earlier, we could subset the corpus before we calculate the tf-idf score to understand which words are most important for a single president within their sotu speeches. Let's do that for Obama. 

```{r}
obama_tf_idf <- tidy_sotu_words %>%
  filter(president == "Barack Obama") %>%
  count(doc_id, word, sort = TRUE) %>%
  bind_tf_idf(word, doc_id, n) %>%
  arrange(desc(tf_idf))

obama_tf_idf
  
```

Based on what you know of the Obama years and sotu speeches generally, how would you interpret these results?

Let's try graphing these results, showing the top tf-idf terms per speech for Obama's speeches. 

```{r}
obama_tf_idf %>%
  group_by(doc_id) %>%
  mutate(word = factor(word, levels = rev(unique(word)))) %>% 
  group_by(doc_id) %>% 
  top_n(5) %>% 
  ungroup() %>%
  ggplot(aes(word, tf_idf, fill = doc_id)) +
  geom_col(show.legend = FALSE) +
  labs(x = NULL, y = "tf-idf") +
  facet_wrap(~doc_id, ncol = 2, scales = "free") +
  coord_flip() + 
  theme(axis.text.y = element_text(angle = 45)) 
```

## N-Grams
https://www.tidytextmining.com/ngrams.html#counting-and-correlating-pairs-of-words-with-the-widyr-package
We have mentioned n-grams in the intro, let's revisit this here and take a look at the most common bigrams in the speeches. Remember this is what we get back:

```{r bigrams-unnest}
sotu_whole %>%
  unnest_tokens(bigram, text, token = "ngrams", n = 2) # create bigram
```

Let's see the most common bigrams:

```{r bigrams-count}
sotu_whole %>%
  unnest_tokens(bigram, text, token = "ngrams", n = 2) %>% 
  count(bigram, sort = TRUE) # count ocurrences and sord descending
```

Ok, so we again need to remove the stopwords. This time let's use dplyr's `filter` function for this. And before that we will `separate` the two words into two columns.

```{r bigrams-remove-stop}
sotu_bigrams <- sotu_whole %>%
  unnest_tokens(bigram, text, token = "ngrams", n = 2) %>% 
  separate(bigram, c("word1", "word2"), sep = " ") %>% # separate into cols
  filter(!word1 %in% stop_words$word) %>% # remove stopwords
  filter(!word2 %in% stop_words$word)

sotu_bigrams %>% 
  count(word1, word2, sort = TRUE)
```

(Bonus question: What happened on that June 30th?)

A bigram can also be treated as a term in a document in the same way that we treated individual words. That means we can look at tf-idf values in the same way. 

First we will re-unite the two word columns again, and then generate the tf-idf count as above.

```{r bigram-tf-idf}
bigram_tf_idf <- sotu_bigrams %>%
  unite(bigram, word1, word2, sep = " ") %>% # combine columns
  count(president, bigram) %>%
  bind_tf_idf(bigram, president, n) %>%
  arrange(desc(tf_idf))
```


What makes the speeches of different presidents unique?

Let's pick a few presidents and plot their highest scoreing tf-idf values here.

```{r bigram-tf-idf-plot}
potus <- c("John F. Kennedy", "Richard M. Nixon", "William J. Clinton", "George Bush", "George W. Bush", "Jimmy Carter")

bigram_tf_idf %>%
  filter(president %in% potus) %>% 
  group_by(president) %>% 
  top_n(20) %>% 
  ggplot(aes(reorder(bigram, tf_idf), tf_idf, fill = president)) +
  geom_col(show.legend = FALSE) +
  labs(x = NULL, y = "tf-idf") +
  facet_wrap(~president, scales = "free", nrow = 2) +
  coord_flip()
```


## Co-ocurrcence

Co-occurances give us a sense of words that appear in the same text, but not necessarily next to each other.

For this section we will make use of the `widyr` package. It allows us to turn our table into a wide matrix. In our case that matrix will be made up of the individual words and the cell values will be the counts of how many times they co-occur. Then we will turn the matrix back into a tidy form, where each row contains the word pairs and the count of their co-occurrence. This lets us count common pairs of words co-appearing within the same speech.

The function which helps us do this is the `pairwise_count()` function. 

Since processing the entire corpus would take too long here, we will only look at the last 20 words of each speech.


```{r pairwise-count}
library(widyr)

# extract last 100 words from text
sotu_whole$speech_end <- word(sotu_whole$text, -100, end = -1)

sotu_word_pairs <- sotu_whole %>% 
  unnest_tokens(word, speech_end) %>% 
  filter(!word %in% stop_words$word) %>% # remove stopwords
  pairwise_count(word, doc_id, sort = TRUE, upper = FALSE) # don't include upper triangle of matrix

```

To plot the co-ocurrcance network, we use the `igraph` library to convert our table into a network graph and `ggraph` which adds functionality to ggplot and makes it easier create a network plot.

```{r plot-network} 
library(igraph)
library(ggraph)

sotu_word_pairs %>% 
  filter(n >= 10) %>%  # only word pairs that occur 10 or more times
  graph_from_data_frame() %>% #convert to graph
  ggraph(layout = "fr") + # place nodes according to the force-directed algorithm of Fruchterman and Reingold
  geom_edge_link(aes(edge_alpha = n, edge_width = n), edge_colour = "tomato") +
  geom_node_point(size = 5) +
  geom_node_text(aes(label = name), repel = TRUE, 
                 point.padding = unit(0.2, "lines")) +
  theme_void()
```

There are alternative approaches for this as well. See for example the `findAssocs` function in the `tm` package.


## Document-Term Matrix

A [document-term matrix (DTM)](https://en.wikipedia.org/wiki/Document-term_matrix) is a format which is frequently used in textanalysis. It is a matrix where we can see the counts of each term per document. In a DTM each row represents a document, each column represents a term, and the cell values are the counts of the ocurrences of the term for the particular document.

`tidytext` provides functionality to convert to and from DTMs, if for example, your analyis requires specific functions that require you to use a different R package which only works with DTM objects.

The `cast_dtm` function can be used to create a DTM object from a tidy table.

Let's assume that for some reason we want to use the `findAssoc` function from the `tm` package.

First we use dplyr to create a table with the document name, the term, and the count.

```{r term-count}
# make a table with document, term, count
tidy_sotu_words %>% 
  count(doc_id, word) 
```

Now we cast it as a DTM.

```{r cast-dtm}
sotu_dtm <- tidy_sotu_words %>% 
  count(doc_id, word) %>% 
  cast_dtm(doc_id, word, n) 

class(sotu_dtm)
```

Finally, let's use it in the `tm` package.

```{r tm-dtm}
library(tm)

# look at the terms with tm function
Terms(sotu_dtm) %>% tail()

# most frequent terms
findFreqTerms(sotu_dtm, lowfreq = 5000)
              
# find terms associated with ...
findAssocs(sotu_dtm, "citizen", corlimit = 0.5)
```

Conversely, `tidytext` implements the `tidy` function (originally from the `broom` package) to import DocumentTermMatrix objects. Note that it only takes the cells from the DTM that are not 0, so there will be no rows with 0 counts.


## Sentiment analysis




TODO: length over time...other similar measures  
TODO: variation between the different presidents?  
TODO: Sentiment analysis and topic modeling ?  

